Locally-Oriented Programming:\\ A Simple Programming Model for Stencil-Based Computations on Multi-Level Distributed Memory Architectures

### 2015-01-27 papers/LOPe/appendix-a.tex
Finished second reading of programming model.
First complete draft of sections 2 and 3.

\section*{Appendix: A Fortran Primer}

This section provides a brief description of Fortran syntax for readers unfamiliar with Fortran syntax.

>>>>

\section*{Appendix: A Fortran Primer for the C Programmer}

This section provides a brief description of Fortran syntax for readers unfamiliar with Fortran
syntax.  The most common source of confusion is array notation.  Fortran uses parentheses
\texttt{()} for both functions arguments and array indexing; thus the confusing expression
\texttt{foo(i,j)} can either be an array reference or a function call depending on context.  Where C
reserves square brackets \texttt{[]} to denote an array reference, Fortran uses square brackets to
reference coarray memory on a remote, distributed memory process (called an image in Fortran).

Otherwise, Fortran syntax is usually straightforward though it may appear strange to the uninitiated
programmer.  A subroutine is a function with no return value and must be called in a statement
rather than as an expression.  Simple syntatic elements are: \texttt{!} the line comment symbol;
\texttt{\&} the line continuation symbol; and \texttt{::} is syntatic sugar used for the separation
of elements, e.g., separating type attributes from the variable list in a type declaration
statement.

### 2015-01-27 papers/LOPe/coarray-comparison.tex
First complete draft of sections 2 and 3.

LOPe provides a purely \emph{local} viewpoint; the programmer is only provided read and write access
to the local array element and write access to a small halo region surrounding the local element.
There is simply \emph{no} mechanism provided for the programmer to even know \emph{where} the local
element is in the context of the broader array.  On a distributed memory architecture, the halo
elements may not even be physically located on the same processor.  If executed on a cluster
containing hybrid processing elements (e.g. GPUs), the halo elements may be as far as three hops
away: one to get to the host processor and another two to get to memory on the hybrid processor
executing on another distributed memory node.  LOPe provides a complete separation between algorithm
development and memory management and synchronization (between memory copies of the same logical
array region covered by halos).  By explicitly describing the existance and size of an array's halo
region, the compiler is provided with enough information to manage most of the hard and detailed work
involved in memory transfer and synchronization.

Additionally, the semantics of the LOPe execution model (on which more will be said later)
remove even the \emph{possibility} of race conditions developing during execution of a concurrent
procedure.

We emphasize some of these advantages by compairing the \texttt{Laplace} implementation shown above
with the implementation of the same algorithm from the original Numrich and Reid paper describing
coarrays in Fortran.  We should point out that this comparison is somewhat unfair because
Numrich and Reid were describing the advantages of coarray notation for transferring memory
on distributed memory architectures, not how ideally to use coarrays within a major code.  But
this example serves to highlight the advantages of LOPe as described.  In the coarray example
shown below type declarations have been removed to save space:
{\small \begin{verbatim}
subroutine Laplace (nrow,ncol,U)
   left = me-1     ! me refers to the current image
   if (me == 1) left = ncol
   right = me + 1
   if (me == ncol) right = 1
   call sync_all( [left,right] ) ! Wait if left and right have
                                 ! not already reached here
   new_u(1:nrow)=new_u(1:nrow)+u(1:nrow)[left]+u(1:nrow)[right]
   call sync_all( [left,right] )
   u(1:nrow) = new_u(1:nrow) - 4.0*u(1:nrow)
end subroutine
\end{verbatim}}

The advantages of LOPe over this example are now described.  Please note the the following
is not a criticism of Coarray Fortran (CAF) as CAF is a general purpose parallel programming
language and LOPe only pertains to the halo pattern useful in stencil-based algorithms.

%%Please note that this example is somewhat unfair because in practice CAF is usually refactored in a locally oriented way by so that communication and synchronization separated into separate procedures.  Locally-oriented programming should be viewed as programming methodology with LOPe as a particular instance.
\subsubsection{LOPe advantages.}
\begin{itemize}

\item
LOPe requires that the implementation of the algorithm to be separate from the call to
effect the halo transfer.  Removing boundary condition specification (e.g., the cyclic
boundary conditions implemented in the CAF example) from the algorithm allows the boundary
conditions to be changed and without changing algorithm code.

\item
LOPe applies the transfer of halo memory across multiple (possibly) levels of memory with
the LOPe intrinsic \texttt{TRANSFER\_HALO} function (described later).  Thus the LOPe
algorithm can be run on a machine with many interconnected nodes, each containing hybrid
processor cores.  The coarray example can only be run on multiple nodes (called images in
Fortran) without accelerator cores.

\item
The algorithm implementation is separate from user-specified synchronization, e.g.,
\texttt{call sync\_all}.  In LOPe, synchronization is subsummed in the semantics of the
\texttt{CONCURRENT} attribute and the \texttt{TRANSFER\_HALO} function described later.

@@

\item
The semantics of LOPe makes explicit management of array temporaries (e.g., \texttt{U} and
\texttt{new\_U} by programmers unnecessary (though still possible).  Because in LOPe the
halo region is a language construct, the compiler is better able to manage temporary
buffers than users on the target hardware platform.

@@

\item
A programmer is not able to store data to the halo region.  If this were allowed, one
thread could overwrite another threads data at undefined times.  The compiler is able to
catch this class of error.

@@

\item
A programmer can't make synchronization errors as synchronization is implicit in
the \texttt{CONCURRENT} attribute.  A thread running a concurrent procedure is provided
with a copy of it's array element (plus halo) that is consistent with the state
of memory at the time of invocation of procedure.  Stores to an individual
thread's local array element (by that thread) is never visible to other threads.
LOPE encourages the creation of small functions and lets the compiler
fuse the procedures together to provide the necessary synchronization.

>>>>

LOPe provides a purely \emph{local} viewpoint; the programmer is only provided read and write access
to the local array element and write access to a small halo region surrounding the local element.
There is simply \emph{no} mechanism provided for the programmer to even know \emph{where} the local
element is in the context of the broader array.  On a distributed memory architecture, the halo
elements may not even be physically located on the same processor.  If executed on a cluster
containing hybrid processing elements (e.g. GPUs), the halo elements may be as far as three hops
away: one to get to the host processor and another two to get to memory on the hybrid processor
executing on another distributed memory node.  LOPe provides a complete separation between algorithm
development and memory management and synchronization (between memory copies of the same logical
array region covered by halos).  By explicitly describing the existance and size of an array's halo
region, the compiler is provided with enough information to manage most of the hard and detailed
work involved in memory transfer and synchronization.  Additionally, the semantics of the LOPe
execution model remove the possibility of race conditions developing during execution of a
concurrent procedure.

We emphasize some of these advantages by comparing the \texttt{Laplace} implementation with the
implementation of the same algorithm from the original Numrich and Reid paper describing coarrays in
Fortran.  We should point out that this comparison is somewhat unfair, because Numrich and Reid were
introducing coarray notation for transferring memory on distributed memory architectures, not
demonstrating how ideally to use coarrays within a large application.
%%Please note that this example is somewhat unfair because in practice CAF is usually refactored in a locally oriented way by so that communication and synchronization separated into separate procedures.  Locally-oriented programming should be viewed as programming methodology with LOPe as a particular instance.
However this example serves
to highlight some of the advantages of LOPe and CAFe as introduced above.  Note that in the coarray
example shown below, type declarations have been removed to save space:
{\small \begin{verbatim}
  subroutine Laplace (nrow,ncol,U)
    left = me-1      ! me refers to the current image
    if (me == 1) left = ncol
    right = me + 1
    if (me == ncol) right = 1
    call sync_all( [left,right] )   ! Wait for left and right
    new_u(1:nrow) = new_u(1:nrow) + u(1:nrow)[left] + u(1:nrow)[right]
    call sync_all( [left,right] )
    u(1:nrow) = new_u(1:nrow) - 4.0*u(1:nrow)
  end subroutine
\end{verbatim}}

The advantages of LOPe and CAFe over this example are now described.  Please note that the following
is not a criticism of coarray Fortran as CAF is a general purpose parallel programming language and
LOPe only pertains to the halo pattern useful in stencil-based algorithms.

\subsubsection{LOPe advantages.}
\begin{itemize}

\item
LOPe requires the implementation of the algorithm to be separate from the call to effect the halo
transfer.  Removing boundary condition specification (e.g., the cyclic boundary conditions
implemented in the CAF example) from the algorithm allows the boundary conditions to be changed
without changing algorithm code.

\item
LOPe applies the transfer of halo memory across possibly multiple levels of memory with the LOPe
intrinsic \texttt{TRANSFER\_HALO} function.  Thus the LOPe algorithm can be run on a machine with
many interconnected nodes, each containing hybrid processor cores.  The coarray example can only be
run on multiple nodes (called images in Fortran) without accelerator cores.

\item
Algorithm implementation is separate from user-specified synchronization, e.g., \texttt{call
  sync\_all}.  In LOPe, synchronization is subsummed in the semantics of the \texttt{CONCURRENT}
attribute and the \texttt{TRANSFER\_HALO}.

@@

\item
The semantics of LOPe makes explicit management of array temporaries (e.g., \texttt{u} and
\texttt{new\_u} by the programmer unnecessary (though still possible).  Because in LOPe the
halo region is a language construct, the compiler is better able to manage temporary
buffers than users on the target hardware platform.

@@

\item
A programmer is not able to store data to the halo region during execution of a LOPe concurrent
function.  If this were allowed, one thread could overwrite another threads data at undefined times.
The compiler is able to catch this class of error.

@@

\item
A programmer can't make synchronization errors as synchronization is implicit in the
\texttt{CONCURRENT} attribute.  A thread running a concurrent procedure is provided with a copy of
it's array element (plus halo) that is consistent with the state of memory \emph{at the time of
  invocation of the procedure}.  Stores to an individual thread's local array element (by that
thread) is never visible to other threads.  LOPE encourages the creation of small functions and lets
the compiler fuse the functions together to improve performance and to provide necessary
synchronization.

### 2015-01-27 papers/LOPe/coarray-extensions.tex
First complete draft of sections 2 and 3.

Consider the \texttt{Laplace} concurrent function whose implementation was provided above.
In this section we demonstrate how this function can be called in a context that allows
full access to all of the enterior elements of the array, as well as array element within
the halo boundary region.  Topics highlighted in this section are: 1. Distributed memory
array allocation; 2. The placement or location(s) (possibly multiple) of the memory
allocated for the distributed array; 3. Remote memory transfer; and 4. Remote execution.
This description is within the context of a Domain Specific Language (DSL) extending
Fortran 2008.  These specific extensions are called CAFe, (Coarray Fortran extensions).

We start with the declaration of an array (to be allocated later),
\begin{verbatim}
  real, allocatable, HALO(1:*:1,1:*:1), dimension(:,:), codimensions(:,:) :: U
\end{verbatim}
with an explicit halo size and with two local dimensions (rank) and two distributed memory
codimensions (corank).  The corank of the array is chosen to be identical to the rank of
the array so that the logical process topology aligns is a way that allows a natural
halo exchange between logically neighboring processes.  For example, if my process
location is \texttt{[pcol,prow]}, then the right-hand halo for my local array \texttt{U}
can be obtained by the assignment \texttt{U(M+1,:) = U(1,:)[pcol-1,prow]} where the size
and cosize of \texttt{U} are given by the allocation statement,
\begin{verbatim}
  allocate(U(0:M+1,0:N+1)[MP,*])
\end{verbatim}

This allocation statement specifies that (given the 1 element halo size provided earlier for
\texttt{U}) the local, left halo is \texttt{U(0,:)}, the right is \texttt{U(M+1,:)},
the bottom is \texttt{U(:,0)} and the top is \texttt{U(:,N+1)}, leaving the enterior
region to be \texttt{U(1:M,1:N)}.  In this allocation statement the total number of
process rows \texttt{NP} can be obtained at runtime and is not a compile time parameter
because the actual number of participating processes (in Fortran called images) is
variable, depending on how many processes are requested at startup.  In this discussion,
it is \emph{assumed} that there are no holes in the logical processor topology, thus
$MP*NP = P$, where $P$ is the total number of participating processes (images).

@@

Once a subimage is obtained, memory on the device may be allocated,
\begin{verbatim}
  if (device /= this_image()) then
    allocate(U[device], HALO_SRC=U)   [[device]]
  end if
\end{verbatim}
There are four points to note regarding this memory allocation: 1. Memory is only
allocated if a subimage has been obtained; 2. The location where memory is allocated is
denoted by the regular coarray notation \texttt{U[device]}; 3. The allocated size and halo
attribute of the new array are assumed from the previously allocated local array
\texttt{U} via the notation \texttt{HALO\_SRC=U} (using \texttt{HALO\_SRC} will also copy
the memory to the device); and finally 4. The allocation itself is \emph{executed} on the
subimage device with the notation \texttt{[[device]]}.

@@

Execution of the Laplace task (defined previously) is done using the Fortran \texttt{do
concurrent} construct as shown below,
\begin{verbatim}
   do while (.not. converged)
      do concurrent (i=1:M, j=1:N)   [[device]]
         call Laplace(U(i,j)[device])
      end do
      call HALO_TRANSFER(U, BC=CYCLIC)
   end do
\end{verbatim}
There are several points that require highlighting: 1. Iteration occurs over the interior
of the array domain, \texttt{(i=1:M, j=1:N)}; 2. Execution of the loop occurs on the
specific subimage indicated by \texttt{[[device]]}; 3. Execution of the iterates may occur
in any order, even \emph{concurrently}; 4. The \texttt{local element} of the array (as
described above in reference to the definition of the \texttt{CONCURRENT} procedure
\texttt{Laplace} is given by the indices \texttt{(i,j)}; 5. Location of memory for the
task is to be taken from the subimage as noted by \texttt{[device]}; and 6. Transfer of
all requisite halo regions is effected by the call to the new LOPe intrinsic function
\texttt{HALO\_TRANSFER()}.

@@

Finally, deallocation of memory is similar to memory allocation:
\begin{verbatim}
   deallocate(U)
   if (device /= this_image()) then
      deallocate(U[device])   [[device]]
   end if
\end{verbatim}

\subsection{Execution Symantics and Memory Management}

This section describes the hierarchical memory layout and how memory consistency between
the 3-levels of memory is maintained.  It describes what is the compiler's responsibility
and what is the programmer's responsibility.

@@

Note that the use of halo cells is the normal way that large and complex MPI and CAF
programs are implemented.  LOPe proposes to formalize this common pattern into the Fortran
language thus providing the compiler information in order to spread computation over more
hardware resources and reduce complexity for the programmer.

>>>>

Consider the \texttt{Laplace} concurrent function discussed above.  In this section we demonstrate
how this function can be called in a context that allows full access to all of the interior elements
of the array, as well as array elements within the logically exterior, halo boundary region.  Topics
highlighted in this section are:
1. Distributed memory array allocation;
2. Explicit memory placement;
3. Remote memory transfer; and
4. Remote execution.
This description is within the context of a Domain Specific Language (DSL), again extending Fortran
2008.  As shorthand, these specific extensions are referred to as CAFe, for Coarray Fortran extensions.

\subsection{CAFe Example}

We start with the declaration of an array (to be allocated later),
\begin{verbatim}
  real, allocatable, HALO(1:*:1,1:*:1),          &
        dimension(:,:), codimensions(:,:) :: U
\end{verbatim}
with an explicit halo size and with two local dimensions (rank) and two distributed memory
codimensions (corank).  The corank of the array is chosen to be identical to the rank of
the array so that the logical process topology aligns is a way that allows a natural
halo exchange between logically neighboring processes.  For example, if my process
location is \texttt{[pcol,prow]}, then the right-hand halo for my local array \texttt{U}
can be obtained by the assignment \texttt{U(M+1,:) = U(1,:)[pcol+1,prow]} where the size
and cosize of \texttt{U} are given by the allocation statement,
\begin{verbatim}
  allocate( U(0:M+1,0:N+1)[MP,*] )
\end{verbatim}

This allocation statement specifies (given the one element halo size provided earlier for
\texttt{U}) that the left halo column is \texttt{U(0,:)}, the right column is \texttt{U(M+1,:)}, the
bottom row is \texttt{U(:,0)} and the top row is \texttt{U(:,N+1)}, leaving the enterior region
\texttt{U(1:M,1:N)}.  In this allocation statement, the total number of process rows \texttt{NP} can
be obtained at runtime, but may not be explicitly provided (according to CAF rules) because the
actual number of participating processes (in Fortran called images) is variable, depending on how
many processes are requested at startup.  In this discussion, it is \emph{assumed} that there are no
holes in the logical processor topology, thus $MP*NP = P$, where $P$ is the total number of
participating processes (images).

@@

Once a subimage is obtained, memory on the device may be allocated,
\begin{verbatim}
  if (device /= this_image()) then
    allocate(U[device], HALO_SRC=U)   [[device]]
  end if
\end{verbatim}
There are four points to note regarding this memory allocation: 1. Memory is only allocated if a
subimage has been obtained; 2. The location where memory is allocated is denoted by regular coarray
notation \texttt{U[device]}; 3. The allocated size and halo attribute of the new array are assumed
from the previously allocated local array \texttt{U} via the notation \texttt{HALO\_SRC=U} (using
\texttt{HALO\_SRC} will also initially copy \texttt{U} to the subimage); and finally 4. The
allocation itself is \emph{executed} on the subimage device with the notation \texttt{[[device]]}.

@@

Execution of the \texttt{Laplace} task (defined previously) is done using the Fortran \texttt{do}
\texttt{concurrent} construct as shown,
\begin{verbatim}
   do while (.not. converged)
      do concurrent (i=1:M, j=1:N)   [[device]]
         call Laplace( U(i,j)[device] )
      end do
      call HALO_TRANSFER(U, BC=CYCLIC)
   end do
\end{verbatim}
There are several points that require highlighting: 1. Iteration occurs over the interior
of the array domain, \texttt{(i=1:M, j=1:N)}; 2. Execution of the loop body occurs on the
specific subimage indicated by \texttt{[[device]]}; 3. Execution of the iterates may occur
in any order, even \emph{concurrently}; 4. The local element of the array (as
defined above in reference to the definition of the concurrent procedure
\texttt{Laplace}) is given by the indices \texttt{(i,j)}; 5. Location of memory for the
task is to be taken from the subimage as noted by \texttt{[device]}; 5. All threads must finish
execution of the loop body before further execution of the program proceeds; and 6. Transfer of
all requisite halo regions is effected by the call to the new LOPe intrinsic function
\texttt{HALO\_TRANSFER()}.  This function is a synchronization event in that all images must
complete the halo transfer before program execution continues.

@@

Finally, memory for the entire array \texttt{U} can be copied from the subimage device
with the statement,
\begin{verbatim}
  U = U[device]
\end{verbatim}
add memory deallocation is similar to memory allocation:
\begin{verbatim}
   deallocate(U)
   if (device /= this_image()) then
      deallocate(U[device])   [[device]]
   end if
\end{verbatim}

%%\subsection{Execution Symantics and Memory Management}

%%This section describes the hierarchical memory layout and how memory consistency between the 3-levels of memory is maintained.  It describes what is the compiler's responsibility and what is the programmer's responsibility.

@@

Note that the use of halo cells is the normal way that large and complex MPI and CAF
programs are implemented.  LOPe proposes to formalize this common pattern into the Fortran
language allowing the compiler access to this information in order to spread computation over more
hardware resources, improve performance, and to reduce complexity for the programmer.
